---
title: "Hydrological model calibration over the historical period"
page-layout: full
---

# Load necessary libraries

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

library(airGR)
library(lubridate)
```

# Load preprocessed observational data

As described in the [**database section**](../loaddata/index.qmd#sec-exportpreprocessed), preprocessed datasets can be efficiently reloaded using the **`readRDS()`** function. In the code chunk below, we reload the historical hydro-climatic dataset that was previously prepared and exported. This dataset serves as the input for the calibration of the hydrological models in the subsequent steps.

```{r}
#| echo: true
#| eval: true

hist_obs = readRDS("output_preprocessed/processed_obsdata_lez.rds")
str(hist_obs)
```


# Calibration of the GR4J model {#sec-gr4jcalib}

The **GR4J** (**`G`**énie **`R`**ural à **`4`** paramètres **`J`**ournalier) model is a parsimonious, lumped conceptual rainfall-runoff model widely used in hydrological studies. It represents catchment processes through a production (soil moisture) reservoir, which controls the transformation of precipitation into effective rainfall by accounting for evapotranspiration and soil water storage, and a routing reservoir, which simulates the delayed transfer of water to the outlet (@fig-gr4jschema). As the name suggests, the model is governed by four parameters: the capacity of the production reservoir ($x_1$), the groundwater exchange coefficient ($x_2$), the routing reservoir capacity ($x_3$), and the time base of the unit hydrograph ($x_4$). For a detailed description of the model structure and parameterization, readers are referred to @PERRIN2003275.

![Schematic structure of the **GR4J rainfall-runoff model** [@PERRIN2003275]. $P$ denotes rainfall (mm), $E$ is potential evapotranspiration (mm), $S$ represents the level of the production reservoir (mm), $R$ the level of the routing reservoir (mm), and $Q$ the simulated streamflow (mm). $UH$ refers to the unit hydrograph, and $F(X₂)$ to the groundwater exchange function.](images/gr4j-illustr.png){#fig-gr4jschema width=55% fig-alt="Schematic structure of the daily GR4J model"}

To optimize the four parameters of the **GR4J model** ($x_1, x_2, x_3, x_4$), we use the **`Michel calibration algorithm`**, integrated within the **`airGR R package`** [@CORON2017166]. The **`Michel algorithm`** is a gradient-free and iterative local search method which begins by exploring the parameter space using a coarse, predefined grid to identify the most promising region of "good" parameter sets. From the best starting point found in the grid, it performs an iterative local search, adjusting the parameters in directions that maximize an objective function, which measures the goodness-of-fit between observed and simulated discharge. Here we use the the **`Kling-Gupta Efficiency (KGE; @GUPTA200980)`** as objective function. 

::: {.callout-note title="Mathematics of the KGE"}
The **KGE** is a decomposition of the **`Nash-Sutcliffe Efficiency (NSE; @NASH1970282)`** designed to provide a more balanced assessment of model performance. Unlike the **`NSE`**, which can disproportionately emphasize high flows, the **KGE** jointly accounts for correlation $(r)$, bias $(\beta)$, and variability $(\alpha)$. **KGE** values range from $-∞$ to $1$, with $1$ indicating perfect agreement with observations.

$$
\text{KGE} = 1 - \sqrt{(r-1)^2 + (\alpha-1)^2 + (\beta-1)^2}
$$

where:

- $r$ is the linear correlation between simulated and observed discharge,
- $\alpha = \frac{\sigma_s}{\sigma_o}$ is the ratio of the standard deviations of simulated ($\sigma_s$) and observed ($\sigma_o$) series,  
- $\beta = \frac{\mu_s}{\mu_o}$ is the ratio of the mean simulated ($\mu_s$) and observed ($\mu_o$) discharge.
:::

The steps described above are adapted from the [**`airGR website`**](https://hydrogr.github.io/airGR/), which provides a comprehensive overview of the package, detailed descriptions of its functions, and practical tutorials to help users get started with hydrological modeling in **`R`**.

::: {.callout-note title="Preparing inputs for **`airGR`** Models"}
To run a hydrological model using the **`airGR`** package, the functions require input data and options in specific formats. To simplify this process, the package provides dedicated functions:

- **`CreateInputsModel()`** – sets up the model inputs such as time series of dates, precipitation, and observed discharge.  
- **`CreateRunOptions()`** – defines model run settings, including warm-up and calibration periods.  
- **`CreateInputsCrit()`** – specifies the efficiency criterion and any transformations of the discharge (e.g., log or square root).  
- **`CreateCalibOptions()`** – configures the calibration algorithm, including which parameters to optimize and predefined values for fixed parameters.

These helper functions ensure that all required inputs are correctly structured for a smooth and error-free model execution.
:::


We first convert the **`Date`** column in the POSIXt format required by the **`airGR ecosystem`**.


```{r}
#| echo: true
#| eval: true

hist_obs$Date = as.POSIXct(hist_obs$Date)
str(hist_obs$Date)
```


Next, we prepare the input data with the `CreateInputsModel()`** function.


```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

InputsModel = CreateInputsModel(
    FUN_MOD = RunModel_GR4J, 
    DatesR = hist_obs$Date,
    Precip = hist_obs$P, 
    PotEvap = hist_obs$PET
)
str(InputsModel)
```


Before running the **GR4J model**, we must select the period for which we want to run the model. It is recommended to include a **warm-up period** of at least one year. This allows the model’s internal reservoirs to reach realistic initial states and helps avoid poor-quality discharge simulations at the beginning of the run caused by inaccurate initialization.


```{r}
#| echo: true
#| eval: true

# define two years of warm-up period
ind_WarmUp = seq(
  which(hist_obs$Date == min(hist_obs$Date)), 
  which(hist_obs$Date == hist_obs$Date[2]+years(2))
)

str(ind_WarmUp)

# run period
ind_Run = seq(
  max(ind_WarmUp) + 1, 
  which(hist_obs$Date == max(hist_obs$Date))
)

str(ind_Run)
```


We can now create the run options using the **`InputsModel`**, **`ind_WarmUp`** and **`ind_Run`** objects as arguments for the **`CreateRunOptions()`** function.


```{r}
#| echo: true
#| eval: true

RunOptions = CreateRunOptions(
  FUN_MOD = RunModel_GR4J,
  InputsModel = InputsModel,
  IndPeriod_WarmUp = ind_WarmUp,
  IndPeriod_Run = ind_Run, 
)
str(RunOptions)
```

Next, we use the **`CreateInputsCrit()`** function that gathers all the data and information required to compute the calibration criterion.

```{r}
#| echo: true
#| eval: true

InputsCrit = CreateInputsCrit(
  # the name of the error criterion functions:
      #* ErrorCrit_RMSE(): Root mean square error (RMSE)
      #* ErrorCrit_NSE(): Nash-Sutcliffe model efficiency coefficient (NSE)
      #* ErrorCrit_KGE(): Kling-Gupta efficiency criterion (KGE)
      #* ErrorCrit_KGE2(): modified Kling-Gupta efficiency criterion (KGE’)
  FUN_CRIT = ErrorCrit_KGE, 
  #* the inputs of the hydrological model previously prepared by the 
  #* CreateInputsModel() function
  InputsModel = InputsModel, 
  #* the options of the hydrological model previously prepared by the 
  #* CreateRunOptions() function
  RunOptions = RunOptions,      
  # the name of the considered variable (by default "Q" for the discharge)
  VarObs = "Q",       
  # observed variable time serie (e.g. the discharge expressed in mm/time step)
  Obs = hist_obs$Q[ind_Run]   
)

str(InputsCrit)
```

Before using the automatic calibration tool, we must prepare the calibration options using the **`CreateCalibOptions()`** function.

```{r}
#| echo: true
#| eval: true

CalibOptions = CreateCalibOptions(  
  FUN_MOD = RunModel_GR4J,        # the name of the model function
  FUN_CALIB = Calibration_Michel  # the name of the calibration algorithm
)

str(CalibOptions)
```

Now we are ready to perform calibration of the GR4J model using the **`Calibration_Michel()`** function.

```{r}
#| echo: true
#| eval: true

OutputsCalib = Calibration_Michel(
  #* the inputs of the hydrological model previously prepared by the 
  #* CreateInputsModel() function
  InputsModel = InputsModel, 
  #* the options of the hydrological model previously prepared by the 
  #* CreateRunOptions() function
  RunOptions = RunOptions,
  #* the options of the hydrological model previously prepared by the 
  #* CreateInputsCrit() function
  InputsCrit = InputsCrit, 
  #* the options of the hydrological calibration model previously prepared by  
  #* the CreateCalibOptions() function
  CalibOptions = CalibOptions,
  #* Le modèle à utiliser 
  FUN_MOD = RunModel_GR4J
)

```

The **`Calibration_Michel()`** function returns a vector containing the optimized parameters of the **GR4J model**. We can extract these values for subsequent use in model simulations and analysis.

```{r}
#| echo: true
#| eval: true

gr4j_params = OutputsCalib$ParamFinalR
kge = OutputsCalib$CritFinal
```

```{r}
#| echo: false

cat(
  " Calibration summary \n", 
  "\tX1: ", gr4j_params[1], "\n",
  "\tX2: ", gr4j_params[2], "\n",
  "\tX1: ", gr4j_params[3], "\n",
  "\tX2: ", gr4j_params[4], "\n",
  "------------------------------\n",
  "\tKGE: ", round(kge, 2), "\n",
  "==============================\n\n"
)
```

Finally, we can vizualize the results of the GR4J model calibration. The **`airGR`** package offers the possibility to make use of the **`plot()`** function.

```{r}
#| echo: true
#| eval: true
#| fig-width: 12
#| fig-height: 8
#| out-width: "100%"
#| label: fig-calibgr4j
#| fig-cap: "GR4J model calibration results: top row shows the total precipitation time serie; second row displays observed versus simulated discharge; third row (bottom-left) presents the interannual average of daily simulated and observed discharge; bottom-middle shows the cumulative frequency plot of simulated and observed discharge; and bottom-right illustrates the correlation between simulated and observed discharge."

OutputsModel = RunModel_GR4J(
  InputsModel = InputsModel, 
  RunOptions = RunOptions, 
  Param = gr4j_params
)
plot(OutputsModel, Qobs = hist_obs$Q[ind_Run])
```


As in the data preprocessing stage, the **calibrated GR4J model** is reused in the subsequent sections. Working directly with the optimized parameters ensures methodological consistency, avoids unnecessary recalibration, and provides a reliable baseline for subsequent analyses and simulations.

```{r}
#| echo: true
#| eval: true

# Create an output directory if it doesn't exist
if (!dir.exists("output_modelcalibration")) dir.create("output_modelcalibration")

# Export using saveRDS
saveRDS(gr4j_params, file="output_modelcalibration/fittedGR4J_lez.rds")
```


# Further Analysis: Calibration of the IHACRES Model {#sec-ihacrescalib}

While the **GR4J model** is used for the remainder of the practical, further analysis may include the calibration of the **IHACRES model**. **IHACRES** stands for **`I`**dentification of unit **`H`**ydrographs **`A`**nd **`C`**omponent flows from **`R`**ainfall, **`E`**vaporation and **`S`**treamflow data. The IHACRES model is a lumped conceptual rainfall-runoff model which architecture is divided into two fundamental modules (@fig-ihacresschema). The **non-linear loss module** converts observed rainfall into effective rainfall by accounting for antecedent moisture conditions through a catchment wetness index, controlled by a small set of parameters governing catchment drying and rainfall–runoff efficiency, typically influenced by temperature or potential evapotranspiration (@fig-ihacresschema). The **linear routing module** then transforms this effective rainfall into discharge using linear reservoirs whose parameters control the response time and relative contribution of quick flow (surface or near-surface runoff) and slow flow (groundwater contribution or baseflow). For a detailed description of the model structure and parameterization, readers are referred to @Jakeman93WR00877.

![Schematic structure of the IHACRES model [@Zhaonh2019047]. $C$ represents mass balance which ensures the long-term volume of effective rainfall matches observed discharge, $\tau_s$ is the reference drying rate, which defines the soil's drying speed at a reference temperature, $f$ is the temperature modulation of drying rate](images/ihacres-illustr.png){#fig-ihacresschema width=100% fig-alt="Schematic structure of the IHACRES model"}

In **`R`**, the most common and robust way to use the **`IHACRES`** model is through the **`hydromad package`**. It is specifically designed for conceptual rainfall-runoff modeling and allows to handle both the **non-linear loss** and **linear routing** modules separately or as a combined model. 

The model requires a time-series object (**`zoo`** or **`xts`**) containing Rainfall ($P$), Temperature ($T$) or Potential Evapotranspiration ($E$), and observed Discharge ($Q$).

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

library(hydromad)

# Convert data into a zoo object
hist_obs$Date = as.Date(hist_obs$Date)
ts_data = with(
  hist_obs, 
  zoo(cbind(P = P, E = PET, Q = Q), order.by = Date)
)

head(ts_data)
```

Next, we define the **IHACRES model** structure by specifying the two modules. In the chunk below, we use the **`CWI (Catchment Wetness Index)`** for the **loss module** and **`expuh (Exponential Unit Hydrograph)`** for the **routing**. 

```{r}
#| echo: true
#| eval: true

# Define the model structure
# sma: Soil Moisture Accounting (Non-linear Loss)
# routing: Linear Routing (Quick and Slow flow)
ihacres = hydromad(ts_data, 
                   sma = "cwi", 
                   routing = "expuh",
                   # Set parameter bounds (example)
                   tau_q = c(0.1, 2),   # Quick flow time constant
                   tau_s = c(10, 100),  # Slow flow time constant
                   v_s = c(0, 1))       # Relative volume of slow flow
```

To calibrate the parameters of the **IHACRES model**, we will use the **Shuffled Complex Evolution (SCE)** global optimizer. The **`SCE algorithm`**, developed at the **University of Arizona [@DUAN1994265]**, is a global optimization strategy specifically designed to efficiently explore high-dimensional and complex parameter spaces where multiple local optima may exist. It is worth noting that the **IHACRES model** involves up to nine (9) parameters to be optimized (cf. @KIM2015543). We first define a robust **KGE** objective function for **`hydromad`**.

```{r}
#| echo: true
#| eval: true
#| code-fold: false
#| code-summary: Show the code

kge_fun = function(Q, X, ...) 
{
  # Remove missing values to prevent cor/sd returning NA
  ok = complete.cases(Q, X)
  if (sum(ok) < 2) return(NA)
  
  Q = Q[ok]
  X = X[ok]
  
  r = cor(X, Q)                     # correlation
  alpha = sd(X) / sd(Q)             # variability ratio
  beta = mean(X) / mean(Q)          # bias ratio
  
  # Calculate KGE
  val = 1 - sqrt((r - 1)^2 + (alpha - 1)^2 + (beta - 1)^2)

  val
}
```


::: {.callout-warning title="Note on calibration time"}
Calibration of the **IHACRES model** using the **`SCE algorithm`** algorithm can be **computationally intensive**. Depending on the **length of the time serie** and the **computer’s processing power**, this step **may take several minutes to complete**. Please be patient and avoid interrupting the process. For testing purposes, the **`control`** list (e.g., `maxit`, `ncomplex`) can be adjusted to reduce runtime.
:::

```{r}
#| echo: true
#| eval: true

# Run calibration using the SCE algorithm and KGE as obj. function
fit_ihacres = fitBySCE(
  ihacres, 
  objective = kge_fun,
  # Run SCE with only 5 iterations (very short, just for testing)
  # Use 2 complexes in the population (again, very minimal).
  control = list(maxit = 5, ncomplex = 2)
)

# Summary of calibrated parameters
summary(fit_ihacres)
```


```{r}
#| echo: true
#| eval: true

#Extract the KGE value
kge_val = objFunVal(fit_ihacres, objective = kge_fun)
cat("IHACRES calibration KGE:", kge_val)
```

Unlike the **`airGR package`**, which provides a dedicated **built-in plotting function** for visualizing model calibration results, the**`hydromad package`** does not include an equivalent comprehensive graphical output. Consequently, we limit the graphical assessment to a direct comparison between observed and simulated discharge time series, which nonetheless provides a clear and informative evaluation of model performance (@fig-calibihacres).

```{r}
#| echo: true
#| eval: true
#| fig-width: 12
#| fig-height: 5
#| label: fig-calibihacres
#| fig-cap: "IHACRES model calibration results: observed versus simulated discharge."

library(lattice)
library(latticeExtra)

xyplot(fit_ihacres,
       with.standard = TRUE,
       xlab = "Time",
       ylab = "Discharge (m3/s)")
```


As for the GR4J model, we can save the calibrated IHACRES model for future use.

```{r}
#| echo: true
#| eval: true

saveRDS(fit_ihacres, file="output_modelcalibration/fittedIHACRES_lez.rds")
```

# References

::: {#refs}
:::