---
title: "Bias correction of climate models"
page-layout: full
---

# Load necessary libraries

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

library(zoo)
library(MBC)
library(dplyr)
library(ggplot2)
library(lubridate)
library(EnvStats)

```

# Load preprocessed data

As described in the [**database section**](../loaddata/index.qmd#sec-exportpreprocessed), preprocessed datasets can be efficiently reloaded using the **`readRDS()`** function. In the code chunk below, we reload the observational and climate projection dataset that was previously prepared and exported. The structure of the **R objects** is fully preserved: the observed data are stored in a dataframe, while the climate projection data are organized as a nested list, ensuring seamless reuse in subsequent analyses.

```{r}
#| echo: true
#| eval: true

hist_obs = readRDS("output_preprocessed/processed_obsdata_lez.rds")
names(hist_obs) = c("Date", "PR", "T", "PET", "Q")

head(hist_obs)
```


Although data have been prepared and stored for three climate models, only the CLM model is used here for demonstration purposes. Readers interested in further exploration may test the other models as an optional exercise.

```{r}
#| echo: true
#| eval: true

clim_proj = readRDS("output_preprocessed/processed_climatedata_lez.rds")
climdataCLM = clim_proj[["climdataCLM"]]

lapply(climdataCLM, function(df) head(df, 5))
```

# Calculate potential evapotranspiration (PET)

Before applying bias correction, we first compute the **PET** from the climate model data. We will use the **`Hargreaves and Samani** equation [@Hargreaves1985], a widely adopted temperature-based method that estimates **PET** from daily temperature extremes and extraterrestrial radiation. Calculating **PET** at this stage ensures that both precipitation and evapotranspiration are consistently represented, which is essential for subsequent hydrological modelling and bias-correction procedures.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: Expand to see the calc_hargreaves(...) function

calc_hargreaves = function(df, lat = 44.2) 
{
  # Convert Latitude to Radians
  phi = lat * pi / 180
  
  # Get Day of Year from the Date column
  doy = yday(df$Date)
  
  # 1. Inverse relative distance Earth-Sun (dr)
  dr = 1 + 0.033 * cos(2 * pi * doy / 365)
  
  # 2. Solar declination (delta)
  delta = 0.409 * sin(2 * pi * doy / 365 - 1.39)
  
  # 3. Sunset hour angle (ws)
  # This handles the geometry of the Earth's curve at 44.2 degrees
  ws = acos(-tan(phi) * tan(delta))
  
  # 4. Extraterrestrial radiation (Ra) in MJ/m2/day
  Ra = (24 * 60 / pi) * 0.0820 * dr * (ws * sin(phi) * sin(delta) + cos(phi) * cos(delta) * sin(ws))
  
  # 5. Convert Ra to evaporation equivalent (mm/day)
  # 1 MJ/m2/day is approx 0.408 mm/day
  Ra_mm = Ra * 0.408
  
  # 6. Final Hargreaves Calculation
  T_mean = (df$TMAX + df$TMIN) / 2
  # pmax ensures we don't get NaNs if Tmin > Tmax (though rare)
  PET = 0.0023 * Ra_mm * (T_mean + 17.8) * sqrt(pmax(0, df$TMAX - df$TMIN))
  
  return(PET)
}
```

```{r}
#| echo: true
#| eval: true

# Apply the calc_hargreaves() function to the 'hist', 'rcp45', and 'rcp85'
climdataCLM = lapply(climdataCLM, 
                      function(df) 
                      {
                        df$PET = calc_hargreaves(df, lat = 44.1958)
                        df[, c("Date", "PR", "PET")]
                      })

lapply(climdataCLM, function(df) head(df, 5))
```

# Bias correction methodology

Bias-correction techniques are commonly applied to reduce systematic discrepancies between climate model simulations and observations, thereby making climate projections suitable for hydrological applications. A wide range of methods has been developed to correct biases in climate variables simulated by climate models.

In this practical, we use the **`MBC package`** to perform bias correction. **`MBC`** stands for **_`M`ultivariate ``B`ias `C`orrection_** and is designed to **calibrate** and **apply multivariate bias-correction algorithms** to climate model simulations involving **multiple climate variables** simultaneously [@Alex2015; @Alex2016; @Alex2018].


## The MBCn algorithm

We will use the **`MBCn` (N-dimensional Multivariate Bias Correction) algorithm** to perform bias correction. The **`MBCn` method** is based on a **probability density function transformation**, making it a powerful approach for correcting biases in climate model outputs (see @Alex2018 for methodological details).

Unlike traditional univariate bias-correction methods that adjust each variable independently, **`MBCn`** corrects multiple variables simultaneously, such as **precipitation** and **potential evapotranspiration (PET)**, while explicitly preserving their **multivariate dependence structure**. This property is particularly important for hydrological modelling, where interactions between climate variables directly influence soil moisture dynamics, evapotranspiration, and runoff generation.

## Accounting for seasonality

To properly account for **seasonality**, the bias-correction procedure is applied **separately for each calendar month**. The data are first split into monthly subsets (e.g., all Januaries together, all Februaries together, and so on). The **`MBCn` algorithm** is then applied independently to each monthly “slice”, after which the corrected data are recombined into continuous time series.

This seasonal stratification ensures that statistical relationships learned for **summer rainfall** are not inadvertently applied to **winter precipitation**, thereby improving the physical consistency and reliability of the bias-corrected climate variables.


# Apply bias correction

The function **`seasonalMBCn(...)`** defined in the chunk below provides a clean wrapper around the **`MBCn()`** routine. It automatically extracts date components, applies the multivariate bias correction to historical and future simulations, and returns bias-corrected datasets that are ready for subsequent hydrological modelling.

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: Show the code


# @param obs: Data frame/matrix of historical observations
# @param mod_hist: Data frame/matrix of historical model outputs
# @param mod_proj: Data frame/matrix of future model projections (RCP)#| code-fold: true
#| code-summary:
# @param var_names: Vector of column names to process
seasonalMBCn = function(obs, 
                        mod_hist, 
                        mod_proj, 
                        var_names = c("PR", "PET")) 
{
  # helper function to add date parts
  add_date_parts = function(df) 
  {
    df$Year  = as.numeric(format(df$Date, "%Y"))
    df$Month = as.numeric(format(df$Date, "%m"))
    df$Day   = as.numeric(format(df$Date, "%d"))

    df
  }

  # add date parts
  obs       = add_date_parts(obs)
  mod_hist  = add_date_parts(mod_hist)
  mod_proj  = add_date_parts(mod_proj)
  
  # Lists to store the corrected chunks for each month
  hist_list = list()
  proj_list = list()
  
  # Loop through each month (1 = Jan, ..., 12 = Dec)
  for (m in 1:12) {

    # Subset data for the current month
    s_obs  = obs[obs$Month == m, ]
    s_hist = mod_hist[mod_hist$Month == m, ]
    s_proj = mod_proj[mod_proj$Month == m, ]
    
    # 2. Convert to matrices for MBCn
    m_obs  = as.matrix(s_obs[, var_names])
    m_hist = as.matrix(s_hist[, var_names])
    m_proj = as.matrix(s_proj[, var_names])
    
    # Run MBCn
    capture.output({ # stop the MBCn from printing the iteration progress
      fit = MBCn(
        o.c = m_obs,
        m.c = m_hist,
        m.p = m_proj,
        iter = 100,
        ratio.seq = c(TRUE, TRUE), 
        trace = .01
      )
    })
    
    # Prepare Adjusted Chunks
    s_hist_adj = as.data.frame(fit$mhat.c)
    colnames(s_hist_adj) = paste0(var_names, "_adj")
    hist_list[[m]] = cbind(s_hist, s_hist_adj)
    
    # Projection
    s_proj_adj = as.data.frame(fit$mhat.p)
    colnames(s_proj_adj) = paste0(var_names, "_adj")
    proj_list[[m]] = cbind(s_proj, s_proj_adj)
  }
  
  # Recombine all months and restore chronological order
  # We use do.call(rbind, ...) to stack the 12 data frames
  final_hist = do.call(rbind, hist_list)
  final_proj = do.call(rbind, proj_list)
  
  # Sort by Year, Month and Dayto keep time series linear
  final_hist = final_hist[
    order(final_hist$Year, final_hist$Month, final_hist$Day), 
  ][,c("Date", "PR", "PR_adj", "PET", "PET_adj")]
  final_proj = final_proj[
    order(final_proj$Year, final_proj$Month, final_proj$Day), 
  ][,c("Date", "PR", "PR_adj", "PET", "PET_adj")]
  
  list(hist = final_hist, proj = final_proj)
}
```

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

# correct RCP45
MBCnSeasonalRCP45 = seasonalMBCn(hist_obs, 
                                 climdataCLM[["hist"]], 
                                 climdataCLM[["rcp45"]])

# correct RCP85
MBCnSeasonalRCP85 = seasonalMBCn(hist_obs, 
                                 climdataCLM[["hist"]], 
                                 climdataCLM[["rcp85"]])

# extract bias corrected data
histCLM_adj = MBCnSeasonalRCP45[['hist']]
rcp45CLM_adj = MBCnSeasonalRCP45[['proj']]
rcp85CLM_adj = MBCnSeasonalRCP85[['proj']]
```
       


# Visualizing bias correction results

To visualize the bias correction results, we plot the the **`Empirical Cumulative Distribution Functions (ECDFs)`** to validate the effectiveness of the **`MBCn algorithm`**.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: Show the code
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| label: fig-biascorr_hist
#| fig-cap: "Bias correction results: Q-Q plots comparing observed precipitation (Precip) and PET with raw and bias-corrected RCM outputs over the historical period."

vars = c("PR", "PET")

par(mfrow = c(1, 2), mar = c(5, 4, 3, 2)) 

for (v in vars) {
  adj_v <- paste0(v, "_adj")

  obs <- na.omit(hist_obs[[v]])
  raw <- na.omit(histCLM_adj[[v]])
  adj <- na.omit(histCLM_adj[[adj_v]])

  if (v == "PR") {
    obs <- obs[obs > 0.1]
    raw <- raw[raw > 0.1]
    adj <- adj[adj > 0.1]
  }

  lims <- range(c(obs, raw, adj))

  qqPlot(x = obs, y = raw,
         plot.type = "Q-Q",
         qq.line.type = "0-1", 
         add.line = TRUE,
         points.col = "#ffb04e",
         line.col = "gray50",
         cex = 1.2,
         pch = 1,
         cex.main = 1.3,
         cex.lab = 1.5,
         cex.axis = 1.4,
         xlim = lims, ylim = lims,
         main = paste("Q-Q Plot:", ifelse(v=="PR", "Precip.", "PET")),
         xlab = "Observations (mm/d)",
         ylab = "RCM data (mm/d)")

  coords_adj <- qqPlot(x = obs, y = adj, plot.it = FALSE)
  points(coords_adj$x, coords_adj$y,
         col = "#5aa4ff",
         pch = 1, cex = 1.2)

  legend("topleft", 
         legend = c("Obs. vs Raw RCM", "Obs. vs  Bias-corrected RCM"),
         col = c("#ffb04e", "#5aa4ff"),
         pch = 1,
         bty = "o",
         cex = 1.3,
         pt.lwd = 2,
         pt.cex = 1.5)
}
```

<!-- To visualize the **climate change signal**, we will now compare the Observed historical data against the Bias-Corrected Future Projections (RCP 4.5 and RCP 8.5). Unlike your previous plot where the green and black lines overlapped perfectly, these plots will show a shift. This shift represents the actual climate change signal (e.g.,different rainfall patterns) after the systematic model biases have been removed. -->

```{r}
#| echo: false
#| eval: false
#| code-fold: false
#| code-summary: Show the code
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| label: fig-biascorr_scenario_fut
#| fig-cap: "Bias correction results: Observed, Raw Model, and Corrected Model distributions for the RCP 4.5 (orange line) and RCP 8.5 (red line) scenarios"

# vars = c("PR", "PET")

# # Use a 1-row, 2-column layout
# par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

# for (v in vars) {

#     # Historical Baseline (Observed)
#     obs_val  = hist_obs[[v]]

#     # raw RCP 4.5 and RCP 8.5
#     raw_rcp45_val = rcp45CLM_adj[[v]]
#     raw_rcp85_val = rcp85CLM_adj[[v]]
      
#     # Corrected Future RCP 4.5 and RCP 8.5
#     adj_rcp45_val = rcp45CLM_adj[[paste0(v, "_adj")]]
#     adj_rcp85_val = rcp85CLM_adj[[paste0(v, "_adj")]]
  
#     # Determine Axis Limits (Including future extremes)
#     if (v == "PR") {
#       plot_log = "x"
#       x_lim = c(0.5, max(c(obs_val, adj_rcp85_val), na.rm = TRUE))
#       x_label = "Precip (mm/day)"
#     } else {
#       plot_log = ""
#       x_lim = c(0.5, max(c(obs_val, adj_rcp85_val), na.rm = TRUE))
#       x_label = "PET (mm/day)"
#     }
    
#     # --- PLOTTING ---
    
#     # Observed Historical (Black - The Baseline)
#     plot(
#       ecdf(obs_val),
#       do.points = FALSE, verticals = TRUE,
#       col = "black", lwd = 1,
#       log = plot_log, xlim = x_lim,
#       main = ifelse(v=="PR", "Precip.", "PET"),
#       xlab = paste(v, "(mm/day)"), ylab = "Probability"
#     )

#     # raw RCP 4.5 (blue) RCP 8.5 (red)
#     plot(
#       ecdf(raw_rcp45_val),
#       add = TRUE,
#       do.points = FALSE, verticals = TRUE,
#       col = "blue", lwd = 1, lty = 2
#     )
    
#     plot(
#       ecdf(raw_rcp85_val),
#       add = TRUE,
#       do.points = FALSE, verticals = TRUE,
#       col = "red", lwd = 1, lty = 2
#     )
    
#     # Corrected RCP 4.5 (green) RCP 8.5 (red)
#     plot(
#       ecdf(adj_rcp45_val),
#       add = TRUE,
#       do.points = FALSE, verticals = TRUE,
#       col = "green", lwd = 1, lty = 1
#     )
    
#     plot(
#       ecdf(adj_rcp85_val),
#       add = TRUE,
#       do.points = FALSE, verticals = TRUE,
#       col = "orange", lwd = 1, lty = 1
#     )
    
#     # Add a legend only to the first plot of each row for clarity
#     if(v == "PR") {
#       legend("bottomright", 
#             legend=c("Obs Hist", 
#                      "Raw RCP 4.5", 
#                      "Raw RCP 8.5", 
#                      "Bias-corrected RCP 4.5", 
#                      "Bias-corrected RCP 8.5"), 
#              col=c("black", "blue", "red", "green", "orange"), lwd = 1.5, bty = "o")
#     }
# }
```

To better understand the seasonal patterns of precipitation and potential evapotranspiration, we compute the mean daily cycle over all years. Unlike a standard calendar year, we align the daily cycle with the hydrological year, starting on September 1st. This allows us to capture the seasonal dynamics of wet and dry periods consistently. To reduce day-to-day variability and highlight the underlying seasonal signal, the resulting daily series are further smoothed using a centered 5-day rolling mean before visualization.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: Show the code
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| label: fig-biascorr_acycle
#| fig-cap: "Seasonal Cycle of Precipitation (top row) and PET (bottom row) over the Hydrological Year: observed (black), raw model (red), and bias-corrected (blue) simulations for historical and future scenarios."

# Helper: compute hydrological day of year (1 = 1 Sept)
hydro_doy = function(dates)
{
  # Shift the year so that 1 Sept = day 1
  doy = yday(dates - months(8))
  return(doy)
}


# Daily climatology based on hydrological year
daily_climatology_hydro = function(df, vars = c("PR_adj", "PET_adj")) 
{
  df %>%
    mutate(HydroDOY = hydro_doy(Date)) %>%
    group_by(HydroDOY) %>%
    summarise(across(all_of(vars), ~mean(.x, na.rm = TRUE))) %>%
    ungroup()
}

roll5 <- function(x) rollmean(x, k = 5, fill = NA, align = "center")

par(mfcol = c(1,2), mar = c(5, 5, 2, 1))

# scenarios = c("Hist", "RCP4.5", "RCP8.5")
scenarios = "Hist"
vars = c("PR", "PET")
cols = c("black", "red", "blue")  # Observed, Raw, Bias-corrected

raw_list = list(
  "Hist"  = histCLM_adj
  # "RCP4.5" = rcp45CLM_adj,
  # "RCP8.5" = rcp85CLM_adj
)

obs_list = list(
  "Hist" = hist_obs
  # "RCP4.5" = hist_obs,
  # "RCP8.5" = hist_obs
)

for(v in vars) {
  for(scn in scenarios) {
    
    # Observed, Raw, Bias-corrected
    obs_val = daily_climatology_hydro(obs_list[[scn]], vars = v)[[v]]
    raw_val = daily_climatology_hydro(raw_list[[scn]], vars = v)[[v]]
    adj_val = daily_climatology_hydro(raw_list[[scn]], vars = paste0(v, "_adj"))[[paste0(v, "_adj")]]
    
    # Apply 5-day rolling mean
    obs_val  = roll5(obs_val)
    raw_val  = roll5(raw_val)
    adj_val  = roll5(adj_val)
    
    # Axis limits
    x_lim = c(1, 365)
    y_lim = range(c(obs_val, raw_val, adj_val), na.rm = TRUE)
    
    # Plot panel
    plot(1:365, obs_val, type="l", lwd=2, col=cols[1], ylim=y_lim,
         xlab="Hydro day of year", 
         ylab=ifelse(v=="PR", "Precip", "PET"),
         main=paste(ifelse(v=="PR", "Precip", "PET"), "-", scn),
         cex.lab = 1.5,   # axis labels
         cex.axis = 1.3,  # axis numbers
         cex.main = 1.5)
    lines(1:365, raw_val, col=cols[2], lwd=1, lty=2)
    lines(1:365, adj_val, col=cols[3], lwd=1, lty=3)
    
    # Legend only for top-left panel (PR-Hist)
    if(v == "PR" && scn == "Hist") {
      legend("topright", legend=c("Observed", "Raw Model", "Bias-corrected"),
             col=cols, lwd=c(2,1.5,1.5), lty=c(1,2,3), bty="o", cex=1.2)
    }
  }
}
```

As in the previous sections, the bias-corrected climate datasets are saved using the **`saveRDS()`** function. This approach preserves the internal structure of the **R objects** (lists and data frames) and facilitates efficient reloading and reuse in subsequent processing steps.

```{r}
#| echo: true
#| eval: true
#| warning: false
#| message: false

# Create an output directory if it doesn't exist
if (!dir.exists("output_biascorrection")) dir.create("output_biascorrection")

# Save the bias-corrected climate data while preserving object structure
saveRDS(list("histCLM_adj" = histCLM_adj,
             "rcp45CLM_adj" = rcp45CLM_adj,
             "rcp85CLM_adj" = rcp85CLM_adj), 
        file="output_biascorrection/biascorrectedCLM.rds")
```


<!-- # What bias correction does NOT do

The “days” used in climate models often do not correspond exactly to the calendar days of the real world. This is not an error, but rather a deliberate mathematical simplification that makes the extensive computations required for climate projections more manageable. As a result, different climate models use different calendars, relying on specialized calendars, such as the **360-day calendar** or the **365-day (no-leap)** calendar. Only a few climate models follow the standard (Gregorian) calendar, with 365 days in non-leap years and 366 days in leap years. Therefore, as done with the data used in this practical, it is recommended to convert climate model output to a standard Gregorian calendar before applying bias correction. -->






<!-- To illustrate this difference, we plot the time series for the first year of the historical period using both the observed data and the climate model data. -->


```{r}
#| echo: false
#| eval: false
#| code-fold: true
#| code-summary: Show the code
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| label: fig-calendarcomp
#| fig-cap: "Time series for the first year of the historical period using both the observed data and the climate model data. The year 2000 is used as a reference year for illustration."

# Reference year
# ref_year <- 2000
# var <- "PR"

# hist_obs$Date    <- as.Date(hist_obs$Date)
# histCLM_adj$Date <- as.Date(histCLM_adj$Date)

# obs_year <- hist_obs[year(hist_obs$Date) == ref_year, ]
# clm_year <- histCLM_adj[year(histCLM_adj$Date) == ref_year, ]

# obs_year$day <- seq_len(nrow(obs_year))
# clm_year$day <- seq_len(nrow(clm_year))

# # Identify the 'leap day' index
# # In 2000 (Leap Year), Feb 29 is index 60.
# leap_day_idx <- 60

# plot(obs_year$day, obs_year[[var]],
#      type = "l", lwd = 1.3,
#      col = "black",
#      cex.lab = 1.5,
#      cex.axis = 1.4,
#      xlab = "Day of year",
#      ylab = "Precipitation (mm/d)",
#      ylim = range(c(obs_year[[var]], clm_year[[var]]), na.rm = TRUE))

# lines(clm_year$day, clm_year[[var]],
#       col = "red", lwd = 1.5, lty = 2)

# lines(clm_year$day, clm_year[[paste0(var, "_adj")]],
#       col = "blue", lwd = 1.5, lty = 3)

# abline(v = leap_day_idx, col = "gray", lty = 4)
# text(leap_day_idx, max(obs_year[[var]]), "Feb 29th", pos = 2, cex = 0.8)

# legend("topleft",
#        legend = c("Observation",
#                   "Raw RCM",
#                   "Bias-corrected RCM"),
#        col = c("black", 
#                "red", 
#                "blue"),
#        lty = c(1, 2, 3),
#        lwd = 3,
#        cex = 1.2,
#        bty = "o")

```




# References

::: {#refs}
:::