---
title: "Bias correction of climate models"
page-layout: full
---

# Load necessary libraries

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

library(zoo)
library(MBC)
library(dplyr)
library(lubridate)
library(EnvStats)
```

# Load preprocessed data

As described in the [**database section**](../loaddata/index.qmd#sec-exportpreprocessed), preprocessed datasets can be efficiently reloaded using the **`readRDS()`** function. In the code chunk below, we reload the observational and climate projection dataset that was previously prepared and exported. The structure of the **R objects** is fully preserved: the observed data are stored in a dataframe, while the climate projection data are organized as a nested list, ensuring seamless reuse in subsequent analyses.

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

hist_obs = readRDS("output_preprocessed/processed_obsdata_lez.rds")
names(hist_obs) = c("Date", "PR", "T", "PET", "Q")

head(hist_obs)
```


Although data have been prepared and stored for three climate models, only the CNRM model is used here for demonstration purposes. Readers interested in further exploration may test the other models as an optional exercise.m_proj = readRDS("output_preprocessed/processed_climatedata_lez.rds")
climdataCNRM = clim_proj[["climdataCNRM"]]

lapply(climdataCNRM, function(df) head(df, 5))

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

clim_proj = readRDS("output_preprocessed/processed_climatedata_lez.rds")
climdataCNRM = clim_proj[["climdataCNRM"]]

lapply(climdataCNRM, function(df) head(df, 5))
```

# Calculate potential evapotranspiration (PET)

Before applying bias correction, we first compute the **PET** from the climate model data. We will use the **`Hargreaves and Samani`** equation [@Hargreaves1985], a widely adopted temperature-based method that estimates **PET** from daily temperature extremes and extraterrestrial radiation. Calculating **PET** at this stage ensures that both precipitation and evapotranspiration are consistently represented, which is essential for subsequent hydrological modelling and bias-correction procedures. To simplify the process, we have provided a ready-to-use function named [**`calc_hargreaves(...)`**](https://github.com/diopBachir/tp_modhydroclimat/blob/main/helper_functions/calc_hargreaves.R) available in the [helper functions folder](https://github.com/diopBachir/tp_modhydroclimat/blob/main/helper_functions/) of this practical’s repository. The chunk below source this function:

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

github_repo = paste0("https://raw.githubusercontent.com/",
                     "diopBachir/tp_modhydroclimat/main/")
source(file.path(github_repo, "helper_functions/calc_hargreaves.R"))
```

```{r}
#| echo: true
#| eval: true

# Apply the calc_hargreaves() function to the 'hist', 'rcp45', and 'rcp85'
climdataCNRM = lapply(climdataCNRM, 
                      function(df) 
                      {
                        df$PET = calc_hargreaves(df, lat = 44.1958)
                        df[, c("Date", "PR", "PET")]
                      })

lapply(climdataCNRM, function(df) head(df, 5))
```


# Bias correction of climate model data

Bias-correction techniques are commonly applied to reduce systematic discrepancies between climate model simulations and observations, thereby making climate projections suitable for hydrological applications. A wide range of methods has been developed to correct biases in climate variables simulated by climate models. In this practical, we will apply the **`MBC method`**. The **_`M`ultivariate `B`ias `C`orrection_** is designed to **calibrate** and **apply multivariate bias-correction algorithms** to climate model simulations involving **multiple climate variables** simultaneously [@Alex2015; @Alex2016; @Alex2018].

Both **`MBC`** and **`CDF-t`** aim to reduce biases in climate model outputs, but they differ in scope and complexity. **`MBC`** handles multiple variables simultaneously, accounting for inter-variable correlations, whereas **`CDF-t`** focuses on single-variable distributions, making it simpler but limited to univariate correction. In practice, the choice of method depends on the variables of interest and the need to preserve multivariate dependence structures in impact studies.


## Applying the MBCn algorithm

We will use the **`MBCn` (N-dimensional Multivariate Bias Correction) algorithm**, from the **`MBC package`**, to perform bias correction. The **`MBCn` method** is based on a **probability density function transformation**, making it a powerful approach for correcting biases in climate model outputs (see @Alex2018 for methodological details).

Unlike traditional univariate bias-correction methods that adjust each variable independently, **`MBCn`** corrects multiple variables simultaneously, such as **precipitation** and **potential evapotranspiration (PET)**, while explicitly preserving their **multivariate dependence structure**. This property is particularly important for hydrological modelling, where interactions between climate variables directly influence soil moisture dynamics, evapotranspiration, and runoff generation.


### Accounting for seasonality

To properly account for **seasonality**, the bias-correction procedure is applied **separately for each calendar month**. The data are first split into monthly subsets (e.g., all Januaries together, all Februaries together, and so on). The **`MBCn` algorithm** is then applied independently to each monthly “slice”, after which the corrected data are recombined into continuous time series.

This seasonal stratification ensures that statistical relationships learned for **summer rainfall** are not inadvertently applied to **winter precipitation**, thereby improving the physical consistency and reliability of the bias-corrected climate variables.


### Apply bias correction

The helper function [**`seasonalMBCn(...)`**](https://github.com/diopBachir/tp_modhydroclimat/blob/main/helper_functions/seasonalMBCn.R) sourced in the chunk below provides a clean wrapper around the **`MBCn()`** routine. It automatically extracts date components, applies the multivariate bias correction to historical and future simulations, and returns bias-corrected datasets that are ready for subsequent hydrological modelling.

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

source(file.path(github_repo, "helper_functions/seasonalMBCn.R"))
```

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false

# correct RCP45
MBCnSeasonalRCP45 = seasonalMBCn(hist_obs, 
                                 climdataCNRM[["hist"]], 
                                 climdataCNRM[["rcp45"]])

# correct RCP85
MBCnSeasonalRCP85 = seasonalMBCn(hist_obs, 
                                 climdataCNRM[["hist"]], 
                                 climdataCNRM[["rcp85"]])

# extract bias corrected data
histCNRM_MBCn_adj = MBCnSeasonalRCP45[['hist']]
rcp45CNRM_MBCn_adj = MBCnSeasonalRCP45[['proj']]
rcp85CNRM_MBCn_adj = MBCnSeasonalRCP85[['proj']]
```
       


### Visualizing bias correction results

To visualize the bias correction results, we plot the the **`Empirical Cumulative Distribution Functions (ECDFs)`** to validate the effectiveness of the **`MBCn algorithm`**.

```{r}
#| echo: true
#| eval: true
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: Show the code
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| label: fig-biascorr_hist
#| fig-cap: "MBCn bias correction results: Q-Q plots comparing observed precipitation (Precip) and PET with raw and bias-corrected RCM using the MBCn algorithm over the historical period."

# variale.s to plot
vars = c("PR", "PET")

# plot layout
par(mfrow = c(1, 2), mar = c(5, 4, 3, 2)) 

for (v in vars) {
  adj_v = paste0(v, "_adj")

  obs = na.omit(hist_obs[[v]])
  raw = na.omit(histCNRM_MBCn_adj[[v]])
  adj = na.omit(histCNRM_MBCn_adj[[adj_v]])

  if (v == "PR") {
    obs = obs[obs > 0.1]
    raw = raw[raw > 0.1]
    adj = adj[adj > 0.1]
  }

  lims = range(c(obs, raw, adj))

  qqPlot(x = obs, y = raw,
         plot.type = "Q-Q",
         qq.line.type = "0-1", 
         add.line = TRUE,
         points.col = "#ffb04e",
         line.col = "gray50",
         cex = 1.2,
         pch = 1,
         cex.main = 1.3,
         cex.lab = 1.5,
         cex.axis = 1.4,
         xlim = lims, ylim = lims,
         main = paste("MBCn - ", ifelse(v=="PR", "Precip.", "PET")),
         xlab = "Observations [mm.d]",
         ylab = "RCM data [mm.d]")

  coords_adj = qqPlot(x = obs, y = adj, plot.it = FALSE)
  points(coords_adj$x, coords_adj$y,
         col = "#5aa4ff",
         pch = 1, cex = 1.2)

  legend("topleft", 
         legend = c("Obs. vs Raw RCM", "Obs. vs  Bias-corrected RCM"),
         col = c("#ffb04e", "#5aa4ff"),
         pch = 1,
         bty = "o",
         cex = 1.3,
         pt.lwd = 2,
         pt.cex = 1.5)
}
```

To better understand the seasonal patterns of precipitation and potential evapotranspiration, we compute the mean daily cycle over all years. Unlike a standard calendar year, we align the daily cycle with the hydrological year, starting on September 1st. This allows us to capture the seasonal dynamics of wet and dry periods consistently. To reduce day-to-day variability and highlight the underlying seasonal signal, the resulting daily series are further smoothed using a centered 30-day rolling mean before visualization.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: Show the code
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| label: fig-biascorr_acycle
#| fig-cap: "Seasonal Cycle of Precipitation (top row) and PET (bottom row) over the Hydrological Year: observed (black), raw model (red), and bias-corrected (blue) simulations for historical and future scenarios. Bias correction is performed using the MBCn algorithm."

# Helper: compute hydrological day of year (1 = 1 Sept)
hydro_doy = function(dates)
{
  # Shift the year so that 1 Sept = day 1
  doy = yday(dates - months(8))
  return(doy)
}


# Daily climatology based on hydrological year
daily_climatology_hydro = function(df, vars = c("PR_adj", "PET_adj")) 
{
  df %>%
    mutate(HydroDOY = hydro_doy(Date)) %>%
    group_by(HydroDOY) %>%
    summarise(across(all_of(vars), ~mean(.x, na.rm = TRUE))) %>%
    ungroup()
}

rollm = function(x) rollmean(x, k = 30, fill = NA, align = "right")

par(mfcol = c(1,2), mar = c(5, 5, 2, 1))

# scenarios = c("Hist", "RCP4.5", "RCP8.5")
scenarios = "Hist"
vars = c("PR", "PET")
cols = c("black", "red", "blue")  # Observed, Raw, Bias-corrected

raw_list = list(
  "Hist"  = histCNRM_MBCn_adj
  # "RCP4.5" = rcp45CNRM_MBCn_adj,
  # "RCP8.5" = rcp85CNRM_MBCn_adj
)

obs_list = list(
  "Hist" = hist_obs
  # "RCP4.5" = hist_obs,
  # "RCP8.5" = hist_obs
)

for(v in vars) {
  for(scn in scenarios) {
    
    # Observed, Raw, Bias-corrected
    obs_val = daily_climatology_hydro(obs_list[[scn]], vars = v)[[v]]
    raw_val = daily_climatology_hydro(raw_list[[scn]], vars = v)[[v]]
    adj_val = daily_climatology_hydro(raw_list[[scn]], vars = paste0(v, "_adj"))[[paste0(v, "_adj")]]
    
    # Apply rolling mean
    obs_val  = rollm(obs_val)
    raw_val  = rollm(raw_val)
    adj_val  = rollm(adj_val)
    
    # Axis limits
    x_lim = c(1, 365)
    y_lim = range(c(obs_val, raw_val, adj_val), na.rm = TRUE)
    
    # Plot panel
    plot(1:365, obs_val, type="l", lwd=2, col=cols[1], ylim=y_lim,
         xlab="Hydro day of year", 
         ylab=ifelse(v=="PR", "Precip", "PET"),
         main=paste(ifelse(v=="PR", "Precip", "PET"), "-", scn),
         cex.lab = 1.5,   # axis labels
         cex.axis = 1.3,  # axis numbers
         cex.main = 1.5)
    lines(1:365, raw_val, col=cols[2], lwd=1, lty=2)
    lines(1:365, adj_val, col=cols[3], lwd=1, lty=3)
    
    # Legend only for top-left panel (PR-Hist)
    if(v == "PR" && scn == "Hist") {
      legend("topright", legend=c("Observed", "Raw Model", "Bias-corrected"),
             col=cols, lwd=c(2,1.5,1.5), lty=c(1,2,3), bty="o", cex=1.2)
    }
  }
}
```

As in the previous sections, the bias-corrected climate datasets are saved using the **`saveRDS()`** function. This approach preserves the internal structure of the **R objects** (lists and data frames) and facilitates efficient reloading and reuse in subsequent processing steps.

```{r}
#| echo: true
#| eval: true
#| warning: false
#| message: false

# Create an output directory if it doesn't exist
if (!dir.exists("output_biascorrection")) dir.create("output_biascorrection")

# Save the bias-corrected climate data while preserving object structure
saveRDS(list("histCNRM_adj" = histCNRM_MBCn_adj,
             "rcp45CNRM_adj" = rcp45CNRM_MBCn_adj,
             "rcp85CNRM_adj" = rcp85CNRM_MBCn_adj), 
        file="output_biascorrection/biascorrectedCNRM-MBCn.rds")
```


## Exercise

Now that we have successfully bias-corrected the CNRM dataset, it is time to apply the same rigor to the second Regional Climate Model (RCM) in our database : the **`SMHI-MPI`** model.


::: {.callout-note title="Hint: load data"}

```{r}
#| echo: true
#| eval: true
#| warning: false
#| message: false

clim_proj = readRDS("output_preprocessed/processed_climatedata_lez.rds")
climdata_SMHI_MPI = clim_proj[["climdataSHMI_MPI"]]

# lapply(climdata_SMHI_MPI, function(df) head(df, 5))
```

::: 


::: {.callout-note title="Hint: Calculate potential evapotranspiration (PET)"}

```{r}
#| echo: true
#| eval: true
#| warning: false
#| message: false

# Apply the calc_hargreaves() function to the 'hist', 'rcp45', and 'rcp85'
climdata_SMHI_MPI = lapply(climdata_SMHI_MPI, 
                           function(df) 
                           {
                              df$PET = calc_hargreaves(df, lat = 44.1958)
                              df[, c("Date", "PR", "PET")]
                           })

# lapply(climdata_SMHI_MPI, function(df) head(df, 5))
```

:::


With the **`SMHI-MPI`** projections loaded and the PET calculated, below are your tasks:

- Execute the Correction: Run the **`seasonalMBCn(...)`** function across your variable matrix (Precipitation and PET).

- Validation: Use the provided visualization suite to generate comparative plots.

- Analysis: Observe how the distribution of the corrected SMHI-MPI data aligns with the reference observations compared to the raw model output.

- Save results: save the bias-corrected data to **`output_biascorrection/biascorrectedSMHI-MPI.rds`**.

::: {.callout-note title="Hint: Target benchmarks"}

Once the **`seasonalMBCn(...)`** function has finished processing, your output visualizations should demonstrate the effectiveness of the correction. Specifically, you are looking for the "collapse" of the model's systematic errors toward the historical observation baseline.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

# correct RCP45
MBCnSeasonalRCP45 = seasonalMBCn(hist_obs, 
                                 climdata_SMHI_MPI[["hist"]], 
                                 climdata_SMHI_MPI[["rcp45"]])

# correct RCP85
MBCnSeasonalRCP85 = seasonalMBCn(hist_obs, 
                                 climdata_SMHI_MPI[["hist"]], 
                                 climdata_SMHI_MPI[["rcp85"]])

# extract bias corrected data
histSMHI_MPI_MBCn_adj = MBCnSeasonalRCP45[['hist']]
rcp45SMHI_MPI_MBCn_adj = MBCnSeasonalRCP45[['proj']]
rcp85SMHI_MPI_MBCn_adj = MBCnSeasonalRCP85[['proj']]
```

```{r}
#| echo: false
#| eval: true
#| code-fold: true
#| code-summary: Show the code
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| label: fig-biascorr_hist_smhi
#| fig-cap: "MBCn bias correction results: Q-Q plots comparing observed precipitation (Precip) and PET with raw and bias-corrected RCM (SMHI-MPI) using the MBCn algorithm over the historical period."

vars = c("PR", "PET")

par(mfrow = c(1, 2), mar = c(5, 4, 3, 2)) 

for (v in vars) {
  adj_v = paste0(v, "_adj")

  obs = na.omit(hist_obs[[v]])
  raw = na.omit(histSMHI_MPI_MBCn_adj[[v]])
  adj = na.omit(histSMHI_MPI_MBCn_adj[[adj_v]])

  if (v == "PR") {
    obs = obs[obs > 0.1]
    raw = raw[raw > 0.1]
    adj = adj[adj > 0.1]
  }

  lims = range(c(obs, raw, adj))

  qqPlot(x = obs, y = raw,
         plot.type = "Q-Q",
         qq.line.type = "0-1", 
         add.line = TRUE,
         points.col = "#ffb04e",
         line.col = "gray50",
         cex = 1.2,
         pch = 1,
         cex.main = 1.3,
         cex.lab = 1.5,
         cex.axis = 1.4,
         xlim = lims, ylim = lims,
         main = paste("MBCn - ", ifelse(v=="PR", "Precip.", "PET")),
         xlab = "Observations [mm.d]",
         ylab = "RCM data [mm.d]")

  coords_adj = qqPlot(x = obs, y = adj, plot.it = FALSE)
  points(coords_adj$x, coords_adj$y,
         col = "#5aa4ff",
         pch = 1, cex = 1.2)

  legend("bottomright", 
         legend = c("Obs. vs Raw RCM", "Obs. vs  Bias-corrected RCM"),
         col = c("#ffb04e", "#5aa4ff"),
         pch = 1,
         bty = "o",
         cex = 1.3,
         pt.lwd = 2,
         pt.cex = 1.5)
}
```


```{r}
#| echo: false
#| eval: true
#| code-fold: true
#| code-summary: Show the code
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 5
#| out-width: "100%"
#| label: fig-biascorr_acycle_smhi
#| fig-cap: "Seasonal Cycle of Precipitation (top row) and PET (bottom row) over the Hydrological Year: observed (black), raw model (red), and bias-corrected (blue) simulations for historical and future scenarios. Bias correction is performed using the MBCn algorithm."

# Helper: compute hydrological day of year (1 = 1 Sept)
hydro_doy = function(dates)
{
  # Shift the year so that 1 Sept = day 1
  doy = yday(dates - months(8))
  return(doy)
}


# Daily climatology based on hydrological year
daily_climatology_hydro = function(df, vars = c("PR_adj", "PET_adj")) 
{
  df %>%
    mutate(HydroDOY = hydro_doy(Date)) %>%
    group_by(HydroDOY) %>%
    summarise(across(all_of(vars), ~mean(.x, na.rm = TRUE))) %>%
    ungroup()
}

rollm = function(x) rollmean(x, k = 30, fill = NA, align = "center")

par(mfcol = c(1,2), mar = c(5, 5, 2, 1))

scenarios = "Hist"
vars = c("PR", "PET")
cols = c("black", "red", "blue")  # Observed, Raw, Bias-corrected

raw_list = list(
  "Hist"  = histSMHI_MPI_MBCn_adj
)

obs_list = list(
  "Hist" = hist_obs
)

for(v in vars) {
  for(scn in scenarios) {
    
    # Observed, Raw, Bias-corrected
    obs_val = daily_climatology_hydro(obs_list[[scn]], vars = v)[[v]]
    raw_val = daily_climatology_hydro(raw_list[[scn]], vars = v)[[v]]
    adj_val = daily_climatology_hydro(raw_list[[scn]], vars = paste0(v, "_adj"))[[paste0(v, "_adj")]]
    
    # Apply rolling mean
    obs_val  = rollm(obs_val)
    raw_val  = rollm(raw_val)
    adj_val  = rollm(adj_val)
    
    # Axis limits
    x_lim = c(1, 365)
    y_lim = range(c(obs_val, raw_val, adj_val), na.rm = TRUE)
    
    # Plot panel
    plot(1:365, obs_val, type="l", lwd=2, col=cols[1], ylim=y_lim,
         xlab="Hydro day of year", 
         ylab=ifelse(v=="PR", "Precip", "PET"),
         main=paste(ifelse(v=="PR", "Precip", "PET"), "-", scn),
         cex.lab = 1.5,   # axis labels
         cex.axis = 1.3,  # axis numbers
         cex.main = 1.5)
    lines(1:365, raw_val, col=cols[2], lwd=1, lty=2)
    lines(1:365, adj_val, col=cols[3], lwd=1, lty=3)
    
    # Legend only for top-left panel (PR-Hist)
    if(v == "PR" && scn == "Hist") {
      legend("topright", legend=c("Observed", "Raw Model", "Bias-corrected"),
             col=cols, lwd=c(2,1.5,1.5), lty=c(1,2,3), bty="o", cex=1.2)
    }
  }
}
```

:::

# What bias correction does NOT do

The **`days`** used in climate models often do not correspond exactly to the **calendar days of the real world**. This is **not an error**, but rather a deliberate **mathematical simplification** that makes the extensive computations required for climate projections more manageable. As a result, different climate models use different calendars, relying on specialized calendars, such as the **`360-day calendar`** or the **`365-day (no-leap)`** calendar. Only a few climate models follow the **standard (Gregorian) calendar**, with **365 days in non-leap years** and **366 days in leap years**. Therefore, as done with the data used in this practical, it is recommended to convert climate model output to a standard Gregorian calendar before applying bias correction.



# References


::: {#refs}
:::