---
title: "Data, Software, Setup and Libraries (Packages)"
page-layout: full
---

# Get the Data Used

The data used in this practical session are available in the project’s dedicated [GitHub repository](https://github.com/diopBachir/tp_modhydroclimat/). The analysis is based on daily observed streamflow records and climate projection data for the `LEZ` watershed. Detailed metadata for this watershed are available on the [HydroPortail website (here)](https://www.hydro.eaufrance.fr/stationhydro/Y320002001/fiche). 

::: {.callout-note icon="false"}
## Download the Data

<a href="https://github.com/diopBachir/tp_modhydroclimat/tree/main/data" download class="download-btn">📥 Download CSV</a>
:::

# R (environment) & Rstudio (IDE)

::: {.callout-note}
An **`IDE (Integrated Development Environment)`** is a **software application** that provides developers with a comprehensive set of tools for **writing, testing, and debugging code** in one place.
:::


## 🧠 What is R?
[**`R`**](https://www.r-project.org/) is a **programming language** and **software environment** designed primarily for **statistical computing**, **data analysis**, and **visualization**.

![Some of the key advantages of learning the R programming language. *(Source: [C# Corner](https://www.c-sharpcorner.com/article/installing-and-setting-up-r-environment-on-windows/))*](img/whylearnR.jpg){#fig-Rbenefits width=100% fig-alt="Benefits of learning R language"}


## 🖥️ What is RStudio?
[**`RStudio`**](https://posit.co/download/rstudio-desktop/) is an **`IDE (Integrated Development Environment)`** specifically made for working with **`R`**. It provides a much more convenient and powerful interface than the basic **`R console`**:

* 📝 Script editor with syntax highlighting and auto-completion
* 💻 Console for interactive execution
* 📦 Environment viewer to see your variables, functions, and data
* 📊 Plot viewer, 🗂️ file browser, 📖 help pane, and more
* ⚙️ Easy access to package management, 🔄 version control, and 📄 [`R Markdown / Quarto`](https://quarto.org/)

![Overview of the four main panes in the **RStudio interface**. **Source pane (top-left):** where scripts and R Markdown files are written and edited. **Console pane (bottom-left):** where R commands are executed interactively. **Environment pane (top-right):** displays the current workspace, including variables and data. **Output pane (bottom-right):** shows plots, files, packages, help pages, and more. *(Source: [Posit](https://docs.posit.co/ide/user/ide/guide/ui/ui-panes.html))*](img/rstudio-panes-labeled.jpeg){#fig-rstudiopane width=100% fig-alt="Rstudio panes labeled"}


## 🔧 How to use them together?

1- Install **`R`** → [https://cran.r-project.org](https://cran.r-project.org)

2- Install **`RStudio`** → [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/)

3- Then, open **`RStudio`** → it will automatically use the installed R engine in the background.

![**R environment**, the engine powering your data analysis, and **RStudio**, the dashboard and controls that make driving (coding) efficient and user-friendly.  *(Source: [umanitoba.ca](https://home.cc.umanitoba.ca/~gersteia/MBIO7040/02-Intro.html))*](img/R-Rstudio.png){#fig-r_rstudio width=100% fig-alt="R and Rstudio"}


# Libraries (Packages)


Below are the **R packages** used in this practical, with links and short descriptions. The simplest, general way to install any R package is:

```{r}
#| echo: true
#| eval: false
#| message: false
#| warning: false

install.packages("package_name") # install from CRAN
library(package_name)            # Load it after installation
```

If the package is on [**GitHub**](https://github.com/), we can use the **`remotes`** or **`devtools`** package:

```{r}
#| echo: true
#| eval: false
#| message: false
#| warning: false

# install remotes if not already installed
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")

# install a package from GitHub
remotes::install_github("username/repo_name")

# Load it
library(package_name)
```

::: {.callout-note title="The R packages used in this practical"}

📦 [**`{dplyr}`**](https://dplyr.tidyverse.org/) — Grammar of data manipulation, providing a consistent set of verbs that solve the most common data manipulation challenge.

📦 [**`{tidyr}`**](https://tidyr.tidyverse.org/) — Help you create tidy data.

📦 [**`{zoo}`**](https://cran.r-project.org/web/packages/zoo/index.html) — Tools and functions for time series analysis and manipulation.

📦 [**`{trend}`**](https://cran.r-project.org/web/packages/trend/index.html) — Provides non-parametric tests test for trend detection.

📦 [**`{lubridate}**](https://cran.r-project.org/web/packages/lubridate/index.html) — To work with date-times and time-spans.

📦 [**`{extRemes}`**](https://www.jstatsoft.org/article/view/v072i08) — For fitting and analyzing extreme value distributions.

📦 [**`{airGR}`**](https://hydrogr.github.io/airGR/) — GR hydrological models toolbox for rainfall–runoff modeling.

📦 [**`{hydromad}`**](https://hydromad.github.io/) — Provide a modelling framework for environmental hydrology.

📦 [**`{EventStats}`**](https://cran.r-project.org/web/packages/EnvStats/index.html) — Graphical and statistical analyses of environmental data.

📦 [**`{MBC}`**](https://github.com/cran/MBC/tree/master) — For calibrating and applying univariate and multivariate bias correction algorithms for climate model simulations of multiple climate variables.

:::