---
title: "Data, Software, Setup and Libraries (Packages)"
page-layout: full
---

# Get the Data Used

The data used in this practical session are available in the project’s dedicated [GitHub repository](https://github.com/diopBachir/tp_modhydroclimat/). The analysis is based on daily observed streamflow records and climate projection data for the `LEZ` watershed. Detailed metadata for this watershed are available on the [HydroPortail website (here)](https://www.sandre.eaufrance.fr/geo/CoursEau_Carthage2017/V52-0400). 

::: {.callout-note icon="false"}
## Download the Data

<a href="https://github.com/diopBachir/tp_modhydroclimat/tree/main/data" download class="download-btn">📥 Download CSV</a>
:::

# R (environment) & Rstudio (IDE)

::: {.callout-note}
An **`IDE (Integrated Development Environment)`** is a **software application** that provides developers with a comprehensive set of tools for **writing, testing, and debugging code** in one place.
:::


## 🧠 What is R?
[**`R`**](https://www.r-project.org/) is a **programming language** and **software environment** designed primarily for **statistical computing**, **data analysis**, and **visualization**.

![Some of the key advantages of learning the R programming language. *(Source: [C# Corner](https://www.c-sharpcorner.com/article/installing-and-setting-up-r-environment-on-windows/))*](img/whylearnR.jpg){#fig-Rbenefits width=100% fig-alt="Benefits of learning R language"}


## 🖥️ What is RStudio?
[**`RStudio`**](https://posit.co/download/rstudio-desktop/) is an **`IDE (Integrated Development Environment)`** specifically made for working with **`R`**. It provides a much more convenient and powerful interface than the basic **`R console`**:

* 📝 Script editor with syntax highlighting and auto-completion
* 💻 Console for interactive execution
* 📦 Environment viewer to see your variables, functions, and data
* 📊 Plot viewer, 🗂️ file browser, 📖 help pane, and more
* ⚙️ Easy access to package management, 🔄 version control, and 📄 [`R Markdown / Quarto`](https://quarto.org/)

![Overview of the four main panes in the **RStudio interface**. **Source pane (top-left):** where scripts and R Markdown files are written and edited. **Console pane (bottom-left):** where R commands are executed interactively. **Environment pane (top-right):** displays the current workspace, including variables and data. **Output pane (bottom-right):** shows plots, files, packages, help pages, and more. *(Source: [Posit](https://docs.posit.co/ide/user/ide/guide/ui/ui-panes.html))*](img/rstudio-panes-labeled.jpeg){#fig-rstudiopane width=100% fig-alt="Rstudio panes labeled"}


## 🔧 How to use them together?

1- Install **`R`** → [https://cran.r-project.org](https://cran.r-project.org)

2- Install **`RStudio`** → [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/)

3- Then, open **`RStudio`** → it will automatically use the installed R engine in the background.

![**R environment**, the engine powering your data analysis, and **RStudio**, the dashboard and controls that make driving (coding) efficient and user-friendly.  *(Source: [umanitoba.ca](https://home.cc.umanitoba.ca/~gersteia/MBIO7040/02-Intro.html))*](img/R-Rstudio.png){#fig-r_rstudio width=100% fig-alt="R and Rstudio"}


# Libraries (Packages)

::: {.callout-important}
## Must-haves
```{r}
#| echo: true
#| eval: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(trend)
library(extRemes)
```
:::

::: {.callout-note}
The packages `dplyr`, `ggplot2` and `lubridate` are part of the [`tidyverse`](https://www.tidyverse.org/packages/) collection of R packages. The `core tidyverse` includes the packages that you’re likely to use in everyday data analyses. It's is advised to install this set of packages, including `dplyr`, `ggplot2` and `lubridate`, together by running:

```{r}
#| echo: true
#| eval: false

install.packages("tidyverse")
```
:::

📦 [{extRemes}](https://www.jstatsoft.org/article/view/v072i08) — For fitting and analyzing extreme value distributions.

```{r}
#| echo: true
#| eval: false

install.packages("extRemes")
```

📦 [{trend}](https://cran.r-project.org/web/packages/trend/index.html) — Provides non-parametric tests test for trend detection.

```{r}
#| echo: true
#| eval: false

install.packages("trend")
```

📦 [{dplyr}](https://cran.r-project.org/web/packages/dplyr/index.html) — For data manipulation and tidy workflows

```{r}
#| echo: true
#| eval: false

install.packages("dplyr")
```

📦 [{ggplot2}](https://cran.r-project.org/web/packages/ggplot2/index.html) — For creating clear and publication-quality plots

```{r}
#| echo: true
#| eval: false

install.packages("ggplot2")
```

📦 [{lubridate}](https://cran.r-project.org/web/packages/lubridate/index.html) — To work with date-times and time-spans.

```{r}
#| echo: true
#| eval: false

install.packages("lubridate")
```